---
title: "Customer Churn Analysis"
author: "Kaustubh Saraf"
date: "4/24/2020"
output: html_document
---


```{r}

set.seed(1357)

library(tidyverse) #used for data manipulation
library(rmarkdown) #used for formatting the markdown file
library(lubridate) #Used for manipulating dates

library(glmnet)
library(caret)
library(rpart) #used for classification trees
library(rpart.plot) #used for plotting the trees
library(ROCR) #used for evaluating tree performance
```


**Importing the dataset**
```{r}
telecom.raw <- read.csv("Cust_Churn_Telecom.csv")
```


**Data Inspection**
```{r}
str(telecom.raw)
```

The `SeniorCitizen` is stored as integer. Post investigating the unique values, its should be stored as factor.
Similarly, `customerID` should not be a factor but stored as a string
```{r}
unique(telecom.raw$SeniorCitizen)
length(unique(telecom.raw$tenure))

telecom.raw$SeniorCitizen <- as.factor(telecom.raw$SeniorCitizen)
telecom.raw$customerID <- as.character(telecom.raw$customerID)
telecom.raw$tenure <- as.numeric(telecom.raw$tenure)
```

**Checking `null` values in all the columns**
```{r}
colSums(is.na(telecom.raw))
```

**Correlation across variables**
```{r}
tel.cor.check <- telecom.raw[!is.na(telecom.raw$TotalCharges),]

corr.matrix <- cor(tel.cor.check[,c('MonthlyCharges','TotalCharges')])
corr.matrix
```
There is high correlation in these two variables. Moving ahead, we will be using `MonthlyCharges` column and drop `TotalCharges`



**The tenure column can be clubbed together to reduce the number of levels.**
```{r}
telecom.raw2 <- telecom.raw %>%
  mutate(tenure.new = as.factor(case_when(tenure >= 0 & tenure < 12 ~ '0-12 Months',
                                   tenure >= 12 & tenure < 24 ~ '12-24 Months',
                                   tenure >= 24 & tenure < 36 ~ '24-36 Months',
                                   tenure >= 36 & tenure < 48 ~ '36-48 Months',
                                   tenure >= 48 & tenure < 60 ~ '48-60 Months',
                                   TRUE ~ '>60 Months'))) %>% 
  select(-c(tenure,TotalCharges))

head(telecom.raw2)
```



**Checking the summary statistics**
```{r}
summary(telecom.raw2)
```

There are entries of `No Internet Connection` in columns like `OnlineSecurity`, `OnlineBackup`, `DeviceProtection`, `TechSupport`, `StreamingTV`, `StreamingMovies`. We will merge them with `No` to reduce unnecessary levels.



```{r}
factorrenames <- names(telecom.raw2[8:14])
  
telecom <- telecom.raw2 %>%
        mutate_at(.vars=factorrenames,
                  .funs=~recode_factor(., `No internet service`="No")) %>%
        mutate_at(.vars="MultipleLines",
                  .funs=~recode_factor(., `No phone service`="No"))
  

str(telecom)
```


Distribution of Monthly Charges
```{r}
hist(telecom$MonthlyCharges, prob = T, col = "grey", breaks = 10, main = "Histogram of Monthly Charges", xlab = "Monthly Charges")
```
* There are high number of customers with monthly charges les sthan $ 30
* At the same time, we can see that the number of customers in the buckets of monthly charges greater tha 70 is high and constant across buckets till $ 110


Outlier in Monthly Charges
```{r}
boxplot(telecom$MonthlyCharges,main = "Boxplot : Monthly Charges")
```
There are no outliers in the monthly charges column


## EDA

**Analyzing Churn Rate across gender**
```{r}
ggplot(data = telecom,aes(gender,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Gender") +
  ggtitle("Churn across Gender") +
  labs(fill = 'Churn')
```
The total customers is slightky higher in the male bucket with slight increment in churned customers

**Analyzing Churn Rate across age of the customer**
```{r}
ggplot(data = telecom,aes(SeniorCitizen,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Senior Citizen Type") +
  ggtitle("Churn across Senior Citizen Variable") +
  labs(fill = 'Churn')
```
The churn rate seems high in Senior Citizens even tough the number of customers in this bucket is pretty low

**Analyzing Churn Rate across the payment methods**
```{r}
ggplot(data = telecom,aes(PaymentMethod,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Payment Method") +
  ggtitle("Churn across Payment Method") +
  labs(fill = 'Churn')
```

* The number of customers in each type of payment method is alsomst similar
* There is high churn in customers with `Electronic Check` payment method


# Modelling

Removing the `customerID` from the dataset
```{r}
telecom <- telecom %>% select(-customerID)
```


Splitting the data into training and test set with 80-20 split
```{r}
sample_index<-sample(nrow(telecom),nrow(telecom)*0.8)
telecom.train<-telecom[sample_index,]
telecom.test<-telecom[-sample_index,]
```


```{r}
library(e1071)
```

```{r}
svm_model<-svm(Churn ~ ., data=telecom.train, type="C-classification", kernel="linear")
```


```{r}
summary(svm_model)
```

```{r}
svm_pred_train <- predict(svm_model, telecom.train)
```

```{r}
mean(svm_pred_train == telecom.train$Churn)
```

```{r}
svm_pred_test <- predict(svm_model,telecom.test)
```


```{r}
mean(svm_pred_test == telecom.test$Churn)
```

# ROC Curve

```{r}
svm_pred_train<- predict(svm_model, telecom.train)
MR.svm<-mean(telecom.train$Churn!=svm_pred_train)

pred <- prediction(as.numeric(svm_pred_train), as.numeric(telecom.train$Churn))
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE)
AUC.svm.insample<-unlist(slot(performance(pred, "auc"), "y.values"))
```

```{r}
svm_pred_test<- predict(svm_model, telecom.test)
MR.svm.test<-mean(telecom.test$Churn!=svm_pred_test)

pred <- prediction(as.numeric(svm_pred_test), as.numeric(telecom.test$Churn))
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE)
AUC.svm.outsample<-unlist(slot(performance(pred, "auc"), "y.values"))
```

```{r}
tune.svm <-svm(Churn ~ ., data=telecom.train, type="C-classification", kernel="linear",cost = 10) 
```

  
```{r}
tune_out <-
tune.svm(as.factor(Churn) ~., data=telecom.train,
type = "C-classification", kernel = "linear",
cost = 10^(-1:2), gamma = c(0.1,1,10), coef0 = c(0.1,1,10))
```



```{r}
tune_out$best.parameters$gamma
```
```{r}
tune_out$best.parameters$coef0
```
```{r}
tune_out$best.parameters$cost
```

```{r}


svm_tune <- svm(Churn ~ ., data=telecom.train, type="C-classification", kernel="linear",cost = 0.1,gamma =1,coef0 = 1) 

```

```{r}
svm_tune_pred <- predict(svm_tune,telecom.train)
```

```{r}
mean(telecom.train$Churn!=svm_tune_pred)
```
### Optimal SVM

```{r}
svm_optimal <- svm(Churn ~ ., data=telecom.train, type="C-classification", kernel="linear",cost = 0.1,gamma =1,coef0 = 1) 
```

```{r}
svm_pred_test_opt<- predict(svm_optimal, telecom.test)
MR.svm.test.opt<-mean(telecom.test$Churn!=svm_pred_test_opt)

pred <- prediction(as.numeric(svm_pred_test_opt), as.numeric(telecom.test$Churn))
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE)
AUC.svm.outsample<-unlist(slot(performance(pred, "auc"), "y.values"))
```

```{r}
MR.svm.test.opt
```


# PCA

```{r}
library(FactoMineR)
library(factoextra)
```

```{r}
telecom_pca <- FAMD(telecom,ncp = 25,graph = FALSE,sup.var = 18)
```


```{r}
fviz_screeplot(telecom_pca, ncp=25)
```

```{r}
get_eigenvalue(telecom_pca)
```

```{r}
telecom_prin_comp <- telecom_pca$ind$contrib[,1:3]
```

```{r}
telecom_pca_train <- as.data.frame(cbind(telecom_prin_comp[sample_index,],as.character(telecom[sample_index,"Churn"])))
```

```{r}
telecom_pca_test <- as.data.frame(cbind(telecom_prin_comp[-sample_index,],as.character(telecom[-sample_index,"Churn"])))
```

```{r}
colnames(telecom_pca_train) <- c("Dim.1","Dim.2","Dim.3","Churn")

telecom_pca_train$Dim.1 <- as.numeric(telecom_pca_train$Dim.1)
telecom_pca_train$Dim.2 <- as.numeric(telecom_pca_train$Dim.2)
telecom_pca_train$Dim.3 <- as.numeric(telecom_pca_train$Dim.3)
```

```{r}
colnames(telecom_pca_test) <- c("Dim.1","Dim.2","Dim.3","Churn")

telecom_pca_test$Dim.1 <- as.numeric(telecom_pca_test$Dim.1)
telecom_pca_test$Dim.2 <- as.numeric(telecom_pca_test$Dim.2)
telecom_pca_test$Dim.3 <- as.numeric(telecom_pca_test$Dim.3)
```



```{r}
head(telecom_pca_train)
```

```{r}
head(telecom_pca_test)
```

# Not Req

```{r}
svm_model_pca<-svm(Churn ~ ., data=telecom_pca_train, type="C-classification", kernel="linear")
```

```{r}
svm_pred_train_pca<- predict(svm_model_pca, telecom_pca_train)
MR.svm.pca<-mean(telecom_pca_train$Churn!=svm_pred_train_pca)
```

```{r}
MR.svm.pca
```



```{r}
svm_pred_test_pca<- predict(svm_model_pca, telecom_pca_test)
MR.svm.pca.test<-mean(telecom_pca_test$Churn!=svm_pred_test_pca)
```
```{r}
MR.svm.pca.test
```


```{r}
str(telecom_pca_train)


```

```{r}
str(telecom_pca_test)
```
```{r}
svm_model_pca_opt<-svm(Churn ~ ., data=telecom_pca_train, type="C-classification", kernel="linear",cost=0.1,gamma = 0.1,coef0 = 0.1)
```

```{r}
svm_pred_train_pca_opt<- predict(svm_model_pca_opt, telecom_pca_train)
MR.svm.pca.opt<-mean(telecom_pca_train$Churn!=svm_pred_train_pca_opt)
```

```{r}
MR.svm.pca.opt
```

