---
title: "Customer Churn Analysis"
author: "Kaustubh Saraf"
date: "4/24/2020"
output: html_document
---


```{r}

set.seed(1357)

library(tidyverse) #used for data manipulation
library(rmarkdown) #used for formatting the markdown file
library(lubridate) #Used for manipulating dates


library(rpart) #used for classification trees
library(rpart.plot) #used for plotting the trees
library(ROCR) #used for evaluating tree performance
```


**Importing the dataset**
```{r}
telecom.raw <- read.csv("Cust_Churn_Telecom.csv")
```


**Data Inspection**
```{r}
str(telecom.raw)
```

The `SeniorCitizen` is stored as integer. Post investigating the unique values, its should be stored as factor.
Similarly, `customerID` should not be a factor but stored as a string
```{r}
unique(telecom.raw$SeniorCitizen)
length(unique(telecom.raw$tenure))

telecom.raw$SeniorCitizen <- as.factor(telecom.raw$SeniorCitizen)
telecom.raw$customerID <- as.character(telecom.raw$customerID)
telecom.raw$tenure <- as.numeric(telecom.raw$tenure)
```

**Checking `null` values in all the columns**
```{r}
colSums(is.na(telecom.raw))
```

**Correlation across variables**
```{r}
tel.cor.check <- telecom.raw[!is.na(telecom.raw$TotalCharges),]

corr.matrix <- cor(tel.cor.check[,c('MonthlyCharges','TotalCharges')])
corr.matrix
```
There is high correlation in these two variables. Moving ahead, we will be using `MonthlyCharges` column and drop `TotalCharges`



**The tenure column can be clubbed together to reduce the number of levels.**
```{r}
telecom <- telecom.raw %>%
  mutate(tenure.new = as.factor(case_when(tenure >= 0 & tenure < 12 ~ '0-12 Months',
                                   tenure >= 12 & tenure < 24 ~ '12-24 Months',
                                   tenure >= 24 & tenure < 36 ~ '24-36 Months',
                                   tenure >= 36 & tenure < 48 ~ '36-48 Months',
                                   tenure >= 48 & tenure < 60 ~ '48-60 Months',
                                   TRUE ~ '>60 Months'))) %>% 
  select(-c(tenure,TotalCharges))

head(telecom)
```



**Checking the summary statistics**
```{r}
summary(telecom)
```

Distribution of Monthly Charges
```{r}
hist(telecom$MonthlyCharges, prob = T, col = "grey", breaks = 10, main = "Histogram of Monthly Charges", xlab = "Monthly Charges")
```
* There are high number of customers with monthly charges les sthan $ 30
* At the same time, we can see that the number of customers in the buckets of monthly charges greater tha 70 is high and constant across buckets till $ 110


Outlier in Monthly Charges
```{r}
boxplot(telecom$MonthlyCharges,main = "Boxplot : Monthly Charges")
```
There are no outliers in the monthly charges column


## EDA

**Analyzing Churn Rate across gender**
```{r}
ggplot(data = telecom,aes(gender,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Gender") +
  ggtitle("Churn across Gender") +
  labs(fill = 'Churn')
```
The total customers is slightky higher in the male bucket with slight increment in churned customers

**Analyzing Churn Rate across age of the customer**
```{r}
ggplot(data = telecom,aes(SeniorCitizen,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Senior Citizen Type") +
  ggtitle("Churn across Senior Citizen Variable") +
  labs(fill = 'Churn')
```
The churn rate seems high in Senior Citizens even tough the number of customers in this bucket is pretty low

**Analyzing Churn Rate across the payment methods**
```{r}
ggplot(data = telecom,aes(PaymentMethod,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Payment Method") +
  ggtitle("Churn across Payment Method") +
  labs(fill = 'Churn')
```

* The number of customers in each type of payment method is alsomst similar
* There is high churn in customers with `Electronic Check` payment method


# Modelling

Changing the `Churn` values to 1 and 0
```{r}
telecom$churn <- ifelse(telecom$Churn=='Yes',1,0)

telecom <- telecom %>% select(-Churn)
```


**Train and Test Datasets**
```{r}
sample_index<-sample(nrow(telecom),nrow(telecom)*0.9)
telecom_train<-telecom[sample_index,]
telecom_test<-telecom[-sample_index,]
```



**Logistic regression**
```{r}

nullmodel<-lm(churn~1, data = telecom_train )

summary(nullmodel)

fullmodel<-lm(Churn~., data = telecom_train )

model_step_both<-step(nullmodel,
                      scope=list(lower=nullmodel,upper=fullmodel),
                      direction = 'forward')
```