---
title: "Customer Churn Analysis"
author: "Kaustubh Saraf"
date: "4/24/2020"
output: html_document
---


```{r}

set.seed(1357)

library(tidyverse) #used for data manipulation
library(rmarkdown) #used for formatting the markdown file
library(lubridate) #Used for manipulating dates

library(glmnet)
library(caret)
library(rpart) #used for classification trees
library(rpart.plot) #used for plotting the trees
library(ROCR) #used for evaluating tree performance
```


**Importing the dataset**
```{r}
telecom.raw <- read.csv("Cust_Churn_Telecom.csv")
```


**Data Inspection**
```{r}
str(telecom.raw)
```

The `SeniorCitizen` is stored as integer. Post investigating the unique values, its should be stored as factor.
Similarly, `customerID` should not be a factor but stored as a string
```{r}
unique(telecom.raw$SeniorCitizen)
length(unique(telecom.raw$tenure))

telecom.raw$SeniorCitizen <- as.factor(telecom.raw$SeniorCitizen)
telecom.raw$customerID <- as.character(telecom.raw$customerID)
telecom.raw$tenure <- as.numeric(telecom.raw$tenure)
```

**Checking `null` values in all the columns**
```{r}
colSums(is.na(telecom.raw))
```

**Correlation across variables**
```{r}
tel.cor.check <- telecom.raw[!is.na(telecom.raw$TotalCharges),]

corr.matrix <- cor(tel.cor.check[,c('MonthlyCharges','TotalCharges')])
corr.matrix
```
There is high correlation in these two variables. Moving ahead, we will be using `MonthlyCharges` column and drop `TotalCharges`



**The tenure column can be clubbed together to reduce the number of levels.**
```{r}
telecom.raw2 <- telecom.raw %>%
  mutate(tenure.new = as.factor(case_when(tenure >= 0 & tenure < 12 ~ '0-12 Months',
                                   tenure >= 12 & tenure < 24 ~ '12-24 Months',
                                   tenure >= 24 & tenure < 36 ~ '24-36 Months',
                                   tenure >= 36 & tenure < 48 ~ '36-48 Months',
                                   tenure >= 48 & tenure < 60 ~ '48-60 Months',
                                   TRUE ~ '>60 Months'))) %>% 
  select(-c(tenure,TotalCharges))

head(telecom.raw2)
```



**Checking the summary statistics**
```{r}
summary(telecom.raw2)
```

There are entries of `No Internet Connection` in columns like `OnlineSecurity`, `OnlineBackup`, `DeviceProtection`, `TechSupport`, `StreamingTV`, `StreamingMovies`. We will merge them with `No` to reduce unnecessary levels.



```{r}
factorrenames <- names(telecom.raw2[8:14])
  
telecom <- telecom.raw2 %>%
        mutate_at(.vars=factorrenames,
                  .funs=~recode_factor(., `No internet service`="No")) %>%
        mutate_at(.vars="MultipleLines",
                  .funs=~recode_factor(., `No phone service`="No"))
  

str(telecom)
```


Distribution of Monthly Charges
```{r}
hist(telecom$MonthlyCharges, prob = T, col = "grey", breaks = 10, main = "Histogram of Monthly Charges", xlab = "Monthly Charges")
```
* There are high number of customers with monthly charges les sthan $ 30
* At the same time, we can see that the number of customers in the buckets of monthly charges greater tha 70 is high and constant across buckets till $ 110


Outlier in Monthly Charges
```{r}
boxplot(telecom$MonthlyCharges,main = "Boxplot : Monthly Charges")
```
There are no outliers in the monthly charges column


## EDA

**Analyzing Churn Rate across gender**
```{r}
ggplot(data = telecom,aes(gender,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Gender") +
  ggtitle("Churn across Gender") +
  labs(fill = 'Churn')
```
The total customers is slightky higher in the male bucket with slight increment in churned customers

**Analyzing Churn Rate across age of the customer**
```{r}
ggplot(data = telecom,aes(SeniorCitizen,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Senior Citizen Type") +
  ggtitle("Churn across Senior Citizen Variable") +
  labs(fill = 'Churn')
```
The churn rate seems high in Senior Citizens even tough the number of customers in this bucket is pretty low

**Analyzing Churn Rate across the payment methods**
```{r}
ggplot(data = telecom,aes(PaymentMethod,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Payment Method") +
  ggtitle("Churn across Payment Method") +
  labs(fill = 'Churn')
```

* The number of customers in each type of payment method is alsomst similar
* There is high churn in customers with `Electronic Check` payment method


# Modelling

Removing the `customerID` from the dataset
```{r}
telecom <- telecom %>% select(-customerID)
```


Splitting the data into training and test set with 80-20 split
```{r}
index <- sample(nrow(telecom))
telecom.subset <- telecom[index,]
row.num <- round(nrow(telecom.subset) * 0.8)

telecom.train <- telecom.subset[1:row.num,]
telecom.test <- telecom.subset[(row.num+1):nrow(telecom.subset),]
```

Defining the model eveluation parameters
```{r}
control <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary, # This uses ROC as model eveluation parameter
  classProbs = TRUE,
  verboseIter = FALSE,
  search = "grid",
  savePredictions = TRUE
)
```

Fitting logistic regression model
```{r}
glm_model.telecom <- train(Churn ~ ., data = telecom.train,
                   method="glm", 
                   trControl = control)

summary(glm_model.telecom)
glm_model.telecom
```



Model Eveluation: Insample
```{r}
pred.glm.train<- predict(glm_model.telecom)
MR.glm<-mean(telecom.train$Churn!=pred.glm.train)

pred <- prediction(as.numeric(pred.glm.train), as.numeric(telecom.train$Churn))
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE)
AUC.glm.insample<-unlist(slot(performance(pred, "auc"), "y.values"))
```

Model Eveluation: Out of sample
```{r}
pred.glm.test<- predict(glm_model.telecom,newdata = telecom.test)
MR.glm.out<-mean(telecom.test$Churn!=pred.glm.test)

pred <- prediction(as.numeric(pred.glm.test), as.numeric(telecom.test$Churn))
perf <- performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE)
AUC.glm.insample<-unlist(slot(performance(pred, "auc"), "y.values"))
```

as.numeric(telecom.train$Churn)


```{r}
glmnet_model <- train(Churn ~ ., data = train,
  metric = "ROC",
  method = "glmnet",
  trControl = control,
  preProcess = c("center","scale")
)

plot(glmnet_model)
```

