---
title: "Customer Churn Analysis"
author: "Pratik Gundawade"
date: "4/24/2020"
output: html_document
---

# Libraries

```{r, warning=FALSE,message=FALSE}

set.seed(1357)

library(tidyverse) #used for data manipulation
library(rmarkdown) #used for formatting the markdown file
library(lubridate) #Used for manipulating dates
library(dplyr) # for data wrangling
library(ggplot2) # for plotting awesome graphics
library(plotly) # for making the graphics interactive
library(randomForest) # for bagging tree
library(glmnet)
library(caret)
library(rpart) #used for classification trees
library(rpart.plot) #used for plotting the trees
library(ROCR) #used for evaluating tree performance
library(ipred)
library(party)
library(xgboost)
library(nnet)
library(naivebayes)
library(rattle) #used for plotting trees
library(FactoMineR) #for PCA
library(factoextra) #for PCA
```

# Reading Dataset

**Importing the dataset**
```{r}
telecom.raw <- read.csv("Cust_Churn_Telecom.csv")
```

# Data Inspection

**Data Inspection**
```{r}
str(telecom.raw)
```

The `SeniorCitizen` is stored as integer. Post investigating the unique values, its should be stored as factor.
Similarly, `customerID` should not be a factor but stored as a string

```{r}
unique(telecom.raw$SeniorCitizen)
length(unique(telecom.raw$tenure))

telecom.raw$SeniorCitizen <- as.factor(telecom.raw$SeniorCitizen)
telecom.raw$customerID <- as.character(telecom.raw$customerID)
telecom.raw$tenure <- as.numeric(telecom.raw$tenure)
```

**Checking `null` values in all the columns**
```{r}
colSums(is.na(telecom.raw))
```

**Correlation across variables**
```{r}
tel.cor.check <- telecom.raw[!is.na(telecom.raw$TotalCharges),]

corr.matrix <- cor(tel.cor.check[,c('MonthlyCharges','TotalCharges')])
corr.matrix
```
There is high correlation in these two variables. Moving ahead, we will be using `MonthlyCharges` column and drop `TotalCharges`



**The tenure column can be clubbed together to reduce the number of levels.**
```{r}
telecom.raw2 <- telecom.raw %>%
  mutate(tenure.new = as.factor(case_when(tenure >= 0 & tenure < 12 ~ '0-12 Months',
                                   tenure >= 12 & tenure < 24 ~ '12-24 Months',
                                   tenure >= 24 & tenure < 36 ~ '24-36 Months',
                                   tenure >= 36 & tenure < 48 ~ '36-48 Months',
                                   tenure >= 48 & tenure < 60 ~ '48-60 Months',
                                   TRUE ~ '>60 Months'))) %>% 
  select(-c(tenure,TotalCharges))

#head(telecom.raw2)
```



**Checking the summary statistics**
```{r}
summary(telecom.raw2)
```

There are entries of `No Internet Connection` in columns like `OnlineSecurity`, `OnlineBackup`, `DeviceProtection`, `TechSupport`, `StreamingTV`, `StreamingMovies`. We will merge them with `No` to reduce unnecessary levels.



```{r}
factorrenames <- names(telecom.raw2[8:14])
  
telecom <- telecom.raw2 %>%
        mutate_at(.vars=factorrenames,
                  .funs=~recode_factor(., `No internet service`="No")) %>%
        mutate_at(.vars="MultipleLines",
                  .funs=~recode_factor(., `No phone service`="No"))
  

#glimpse(telecom)
```


Distribution of Monthly Charges
```{r}
hist(telecom$MonthlyCharges, prob = T, col = "grey", breaks = 10, main = "Histogram of Monthly Charges", xlab = "Monthly Charges")
```
* There are high number of customers with monthly charges les sthan $ 30
* At the same time, we can see that the number of customers in the buckets of monthly charges greater tha 70 is high and constant across buckets till $ 110


Outlier in Monthly Charges
```{r}
boxplot(telecom$MonthlyCharges,main = "Boxplot : Monthly Charges")
```
There are no outliers in the monthly charges column


# EDA

**Analyzing Churn Rate across gender**
```{r}
ggplot(data = telecom,aes(gender,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Gender") +
  ggtitle("Churn across Gender") +
  labs(fill = 'Churn')
```
The total customers is slightky higher in the male bucket with slight increment in churned customers

**Analyzing Churn Rate across age of the customer**
```{r}
ggplot(data = telecom,aes(SeniorCitizen,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Senior Citizen Type") +
  ggtitle("Churn across Senior Citizen Variable") +
  labs(fill = 'Churn')
```
The churn rate seems high in Senior Citizens even tough the number of customers in this bucket is pretty low

**Analyzing Churn Rate across the payment methods**
```{r}
ggplot(data = telecom,aes(PaymentMethod,fill = (Churn))) +
  geom_bar(position = 'dodge') +
  ylab("Number of Records") +
  xlab("Payment Method") +
  ggtitle("Churn across Payment Method") +
  labs(fill = 'Churn')
```

* The number of customers in each type of payment method is alsomst similar
* There is high churn in customers with `Electronic Check` payment method


# Models without PCA

Removing the `customerID` from the dataset
```{r}
telecom <- telecom %>% select(-customerID)
```


Splitting the data into training and test set with 80-20 split
```{r}
set.seed(1357)
sample_index<-sample(nrow(telecom),nrow(telecom)*0.8)
telecom.train<-telecom[sample_index,]
telecom.test<-telecom[-sample_index,]
```

## Defining the model evaluation parameters
```{r}
control <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary, # This uses ROC as model eveluation parameter
  classProbs = TRUE,
  verboseIter = FALSE,
  search = "grid",
  savePredictions = TRUE
)
```

```{r}
control1 <- trainControl(
  method = "oob",
  number = 5,
  summaryFunction = twoClassSummary, # This uses ROC as model eveluation parameter
  classProbs = TRUE,
  verboseIter = FALSE,
  search = "grid",
  savePredictions = TRUE
)
```

## Fitting bagging tree model


```{r}
bagg_model.telecom <- train(Churn ~ ., data = telecom.train,
                   method='treebag', 
                   trControl = control1,
                   keepX = T)

bagg_pred <- predict(bagg_model.telecom, newdata = telecom.test)

```
  
```{r}

baggcm <- confusionMatrix(bagg_pred, telecom.test[["Churn"]])
baggcm

```

```{r}
baggaccuracy <- baggcm$overall[c(1,3,4)]
baggaccuracy
```


```{r}
pred.bagg.train<- predict(bagg_model.telecom, newdata = telecom.train)
MR.bagg.in <-mean((telecom.train$Churn != pred.bagg.train)*1)

pred.bagg.test<- predict(bagg_model.telecom, newdata = telecom.test)
MR.bagg.out <-mean((telecom.test$Churn != pred.bagg.test)*1)


MR.bagg.in;
MR.bagg.out
```

### ROC Curve for the bagging model

```{r}
pred_bagg <- prediction(predictions = as.numeric(pred.bagg.test), as.numeric(telecom.test$Churn))

perf_bagg <- performance(pred_bagg,"tpr","fpr")

plot(perf_bagg,colorize=TRUE)

```

**Area Under the curve**

```{r}
unlist(slot(performance(pred_bagg, 'auc'), "y.values"))
```

## Fitting Random Forest tree model

```{r}
rf_model.telecom <- train(Churn ~ ., data = telecom.train,
                   method='rf', 
                   trControl = control1)

rf_pred <- predict(rf_model.telecom, newdata = telecom.test)

rfcm <- confusionMatrix(rf_pred, telecom.test[["Churn"]])
rfcm

```

```{r}
rfaccuracy <- rfcm$overall[c(1,3,4)]
rfaccuracy
```

```{r}
pred.rf.train<- predict(rf_model.telecom)
MR.rf.in <-mean((telecom.train$Churn != pred.rf.train)*1)

pred.rf.test <- predict(rf_model.telecom, newdata = telecom.test)
MR.rf.out <- mean((telecom.test$Churn != pred.rf.test)*1)


MR.rf.in;
MR.rf.out
```

### ROC Curve for the bagging model
```{r}

pred_rf <- prediction(predictions = as.numeric(pred.rf.test), as.numeric(telecom.test$Churn))

perf_rf <- performance(pred_rf,"tpr","fpr")

plot(perf_rf,colorize=TRUE)

```

**Area Under the curve**

```{r}
unlist(slot(performance(pred_rf, 'auc'), "y.values"))
```

## Fitting Gradient Boosting

```{r}
gb_model.telecom <- train(Churn ~ ., data = telecom.train,
                   method='xgbTree',
                   trControl = control)


gb_pred <- predict(gb_model.telecom, newdata = telecom.test)

gbcm <- confusionMatrix(gb_pred, telecom.test[["Churn"]])
gbcm


```

```{r}
gbaccuracy <- gbcm$overall[c(1,3,4)]
gbaccuracy
```

```{r}

pred.gb.train<- predict(gb_model.telecom, newdata = telecom.train)
MR.gb.in <-mean((telecom.train$Churn != pred.gb.train)*1)

pred.gb.test<- predict(gb_model.telecom, newdata = telecom.test)
MR.gb.out <- mean((telecom.test$Churn != pred.gb.test)*1)


MR.gb.in;
MR.gb.out
```

### ROC Curve for the gradient boosting model

```{r}

pred_gb <- prediction(predictions = as.numeric(pred.gb.test), as.numeric(telecom.test$Churn))

perf_gb <- performance(pred_gb,"tpr","fpr")

plot(perf_gb,colorize=TRUE)

```

**Area Under the curve**
```{r}
unlist(slot(performance(pred_gb, 'auc'), "y.values"))
```

## Naive Bayes

```{r, warning=FALSE, message=FALSE}

nb_model.telecom <- train(Churn ~ ., data = telecom.train,
                    method='naive_bayes',
                    trControl = control)
nb_pred <- predict(nb_model.telecom, newdata = telecom.test)

nbcm <- confusionMatrix(nb_pred, telecom.test[["Churn"]])
nbcm
#summary(bagg_model.telecom)
#bagg_model.telecom

```

```{r}
pred.nb.train<- predict(nb_model.telecom)
MR.nb.in <-mean((telecom.train$Churn != pred.nb.train)*1)

pred.nb.test<- predict(gb_model.telecom, newdata = telecom.test)
MR.nb.out <- mean((telecom.test$Churn != pred.nb.test)*1)


MR.nb.in;
MR.nb.out

```
### ROC Curve for the gradient boosting model

```{r}

pred_nb <- prediction(predictions = as.numeric(pred.nb.test), as.numeric(telecom.test$Churn))

perf_nb <- performance(pred_nb,"tpr","fpr")

plot(perf_nb,colorize=TRUE)

```

**Area Under the curve**
```{r}
unlist(slot(performance(pred_nb, 'auc'), "y.values"))
```



# Models With PCA
## PCA

```{r}
telecom_pca <- FAMD(telecom,ncp = 25,graph = FALSE,sup.var = 18)
```


```{r}
fviz_screeplot(telecom_pca, ncp=25)
```

```{r}
get_eigenvalue(telecom_pca)
```

```{r}
telecom_prin_comp <- telecom_pca$ind$contrib[,1:15]
```

```{r}
telecom_pca_train <- as.data.frame(cbind(telecom_prin_comp[sample_index,],as.character(telecom[sample_index,"Churn"])))
```

```{r}
telecom_pca_test <- as.data.frame(cbind(telecom_prin_comp[-sample_index,],as.character(telecom[-sample_index,"Churn"])))
```

```{r}
colnames(telecom_pca_train) <- c("Dim.1","Dim.2","Dim.3","Dim.4","Dim.5","Dim.6","Dim.7",
                                 "Dim.8","Dim.9","Dim.10","Dim.11","Dim.12","Dim.13","Dim.14","Dim.15",
                                 "Churn")

telecom_pca_train$Dim.1 <- as.numeric(telecom_pca_train$Dim.1)
telecom_pca_train$Dim.2 <- as.numeric(telecom_pca_train$Dim.2)
telecom_pca_train$Dim.3 <- as.numeric(telecom_pca_train$Dim.3)
telecom_pca_train$Dim.4 <- as.numeric(telecom_pca_train$Dim.4)
telecom_pca_train$Dim.5 <- as.numeric(telecom_pca_train$Dim.5)
telecom_pca_train$Dim.6 <- as.numeric(telecom_pca_train$Dim.6)
telecom_pca_train$Dim.7 <- as.numeric(telecom_pca_train$Dim.7)
telecom_pca_train$Dim.8 <- as.numeric(telecom_pca_train$Dim.8)
telecom_pca_train$Dim.9 <- as.numeric(telecom_pca_train$Dim.9)
telecom_pca_train$Dim.10 <- as.numeric(telecom_pca_train$Dim.10)
telecom_pca_train$Dim.11 <- as.numeric(telecom_pca_train$Dim.11)
telecom_pca_train$Dim.12 <- as.numeric(telecom_pca_train$Dim.12)
telecom_pca_train$Dim.13 <- as.numeric(telecom_pca_train$Dim.13)
telecom_pca_train$Dim.14 <- as.numeric(telecom_pca_train$Dim.14)
telecom_pca_train$Dim.15 <- as.numeric(telecom_pca_train$Dim.15)

```

```{r}
colnames(telecom_pca_test) <- c("Dim.1","Dim.2","Dim.3","Dim.4","Dim.5","Dim.6","Dim.7",
                                "Dim.8","Dim.9","Dim.10","Dim.11","Dim.12","Dim.13","Dim.14","Dim.15",
                                "Churn")

telecom_pca_test$Dim.1 <- as.numeric(telecom_pca_test$Dim.1)
telecom_pca_test$Dim.2 <- as.numeric(telecom_pca_test$Dim.2)
telecom_pca_test$Dim.3 <- as.numeric(telecom_pca_test$Dim.3)
telecom_pca_test$Dim.4 <- as.numeric(telecom_pca_test$Dim.4)
telecom_pca_test$Dim.5 <- as.numeric(telecom_pca_test$Dim.5)
telecom_pca_test$Dim.6 <- as.numeric(telecom_pca_test$Dim.6)
telecom_pca_test$Dim.7 <- as.numeric(telecom_pca_test$Dim.7)
telecom_pca_test$Dim.8 <- as.numeric(telecom_pca_test$Dim.8)
telecom_pca_test$Dim.9 <- as.numeric(telecom_pca_test$Dim.9)
telecom_pca_test$Dim.10 <- as.numeric(telecom_pca_test$Dim.10)
telecom_pca_test$Dim.11 <- as.numeric(telecom_pca_test$Dim.11)
telecom_pca_test$Dim.12 <- as.numeric(telecom_pca_test$Dim.12)
telecom_pca_test$Dim.13 <- as.numeric(telecom_pca_test$Dim.13)
telecom_pca_test$Dim.14 <- as.numeric(telecom_pca_test$Dim.14)
telecom_pca_test$Dim.15 <- as.numeric(telecom_pca_test$Dim.15)
```

```{r}
#head(telecom_pca_train)
```

```{r}
#head(telecom_pca_test)
```



## Fitting bagging tree model (PCA)


```{r}
pca_bagg_model.telecom <- train(Churn ~ ., data = telecom_pca_train,
                   method='treebag', 
                   trControl = control1,
                   keepX = T)

pca_bagg_pred <- predict(pca_bagg_model.telecom, newdata = telecom_pca_test)

#fancyRpartPlot(bagg_model.telecom$finalModel)

```
  
```{r}

pca_baggcm <- confusionMatrix(pca_bagg_pred, telecom_pca_test[["Churn"]])
pca_baggcm
#summary(bagg_model.telecom)
#bagg_model.telecom

```

```{r}
pca_baggaccuracy <- pca_baggcm$overall[c(1,3,4)]
pca_baggaccuracy
```


```{r}
pca_pred.bagg.train<- predict(pca_bagg_model.telecom, newdata = telecom_pca_train)
pca_MR.bagg.in <-mean((telecom_pca_train$Churn != pca_pred.bagg.train)*1)

pca_pred.bagg.test<- predict(pca_bagg_model.telecom, newdata = telecom_pca_test)
pca_MR.bagg.out <-mean((telecom_pca_test$Churn != pca_pred.bagg.test)*1)


pca_MR.bagg.in;
pca_MR.bagg.out
```

### ROC Curve for the bagging model

```{r}

pca_pred_bagg <- prediction(predictions = as.numeric(pca_pred.bagg.test), as.numeric(telecom_pca_test$Churn))

pca_perf_bagg <- performance(pca_pred_bagg,"tpr","fpr")

plot(pca_perf_bagg,colorize=TRUE)

```

**Area Under the curve**

```{r}
unlist(slot(performance(pred_bagg, 'auc'), "y.values"))
```

## Fitting Random Forest tree model (PCA)

```{r}
pca_rf_model.telecom <- train(Churn ~ ., data = telecom_pca_train,
                   method='rf', 
                   trControl = control1)

pca_rf_pred <- predict(pca_rf_model.telecom, newdata = telecom_pca_test)

pca_rfcm <- confusionMatrix(pca_rf_pred, telecom.test[["Churn"]])
pca_rfcm


```

```{r}
pca_rfaccuracy <- pca_rfcm$overall[c(1,3,4)]
pca_rfaccuracy
```

```{r}
pca_pred.rf.train<- predict(pca_rf_model.telecom, newdata = telecom_pca_train)
pca_MR.rf.in <-mean((telecom_pca_train$Churn != pca_pred.rf.train)*1)

pca_pred.rf.test <- predict(pca_rf_model.telecom, newdata = telecom_pca_test)
pca_MR.rf.out <- mean((telecom_pca_test$Churn != pca_pred.rf.test)*1)


pca_MR.rf.in;
pca_MR.rf.out
```

### ROC Curve for the RandonForest model
```{r}

pca_pred_rf <- prediction(predictions = as.numeric(pca_pred.rf.test), as.numeric(telecom_pca_test$Churn))

pca_perf_rf <- performance(pca_pred_rf,"tpr","fpr")

plot(pca_perf_rf,colorize=TRUE)

```

**Area Under the curve**

```{r}
unlist(slot(performance(pca_pred_rf, 'auc'), "y.values"))
```

## Fitting Gradient Boosting (PCA)

```{r}
pca_gb_model.telecom <- train(Churn ~ ., data = telecom_pca_train,
                   method='xgbTree',
                   trControl = control)


pca_gb_pred <- predict(pca_gb_model.telecom, newdata = telecom_pca_test)

pca_gbcm <- confusionMatrix(pca_gb_pred, telecom_pca_test[["Churn"]])
pca_gbcm


```

```{r}
pca_gbaccuracy <- pca_gbcm$overall[c(1,3,4)]
pca_gbaccuracy
```

```{r}

pca_pred.gb.train<- predict(pca_gb_model.telecom, newdata = telecom_pca_train)
pca_MR.gb.in <-mean((telecom_pca_train$Churn != pca_pred.gb.train)*1)

pca_pred.gb.test<- predict(pca_gb_model.telecom, newdata = telecom_pca_test)
pca_MR.gb.out <- mean((telecom_pca_test$Churn != pca_pred.gb.test)*1)


pca_MR.gb.in;
pca_MR.gb.out
```

### ROC Curve for the gradient boosting model

```{r}

pca_pred_gb <- prediction(predictions = as.numeric(pca_pred.gb.test), as.numeric(telecom_pca_test$Churn))

pca_perf_gb <- performance(pca_pred_gb,"tpr","fpr")

plot(pca_perf_gb,colorize=TRUE)

```

**Area Under the curve**
```{r}
unlist(slot(performance(pca_pred_gb, 'auc'), "y.values"))
```



## Naive Bayes (PCA)

```{r, warning=FALSE,message=FALSE}

pca_nb_model.telecom <- train(Churn ~ ., data = telecom_pca_train,
                    method='naive_bayes',
                    trControl = control)
pca_nb_pred <- predict(pca_nb_model.telecom, newdata = telecom_pca_test)

pca_nbcm <- confusionMatrix(pca_nb_pred, telecom_pca_test[["Churn"]])
pca_nbcm


```

```{r}
pca_pred.nb.train<- predict(pca_nb_model.telecom, newdata = telecom_pca_train)
pca_MR.nb.in <-mean((telecom_pca_train$Churn != pca_pred.nb.train)*1)

pca_pred.nb.test<- predict(pca_gb_model.telecom, newdata = telecom_pca_test)
pca_MR.nb.out <- mean((telecom_pca_test$Churn != pca_pred.nb.test)*1)


pca_MR.nb.in;
pca_MR.nb.out

```
### ROC Curve for the gradient boosting model

```{r}

pca_pred_nb <- prediction(predictions = as.numeric(pca_pred.nb.test), as.numeric(telecom_pca_test$Churn))

pca_perf_nb <- performance(pca_pred_nb,"tpr","fpr")

plot(pca_perf_nb,colorize=TRUE)

```
